Approximation as Differentiation

### 2025-05-06 appendix/notes/auto-diff.tex
Use const family notation.
Experiment with diagram.
Diff* is (I think -- haven't checked) a full subcategory of Fam(FinVect^op).
Cleanup.

\begin{definition}[Forward mode automatic differentiation functor]
Define the functor $\tangents$ which sends every vector space $\RR^n$ to itself and every differentiable map
$f: \RR^m \to \RR^n$ to the function $\tangents(f) = x \mapsto (f(x), \pushf{f}_x): \RR^m \to \RR^n \times
(\RR^m \linearto \RR^n)$, associating to every point its image in $f$ and the forward derivative of $f$ at
that point.
\end{definition}

@@

\begin{definition}[Reverse mode automatic differentiation functor]
Define the functor $\cotangents$ which also sends every vector space $\RR^n$ to itself but which sends every
differentiable map $f: \RR^m \to \RR^n$ to the function $\cotangents(f) = x \mapsto (f(x), \pullf{f}_x): \RR^m
\to \RR^n \times (\RR^n \linearto \RR^m)$.
\end{definition}

\noindent and again we can define a corresponding category $\Diff^*$ of such maps where composition of the
cotangent maps respects the (backward) chain rule.

\subsubsection{Automatic differentiation via category of families}

Recall that the category $\Fam(\FinVect)$ is the Grothendieck construction $\Grothendieck{X}\Fam(X,
\FinVect)$. The functor $\Diff \hookrightarrow \Fam(\FinVect)$ which send objects $\RR^n$ to constant families
of tangent spaces $(\RR^n, \const(\RR^n))$ and morphisms $f: \RR^m \to \RR^n \times (\RR^m \linearto
\RR^n)$ to pairs $(f_1, f_2)$ of differentiable functions $f_1: \RR^m \to \RR^n$ and families of linear maps
$f_2: \{{f_2}_x: \RR^m \linearto \RR^n\}_{x \in \RR^m}$ exhibits $\Diff$ as equivalent to a full subcategory
of $\Fam(\FinVect)$.

For any $(f, \partial f): (\RR^m, \const(\RR^m)) \to (\RR^n, \const(\RR^n))$ and $(g, \partial g): (\RR^n,
\const(\RR^n)) \to (\RR^k, \const(\RR^k))$ in that subcategory, we have that for any $x \in \RR^m$:
\begin{itemize}
\item $\partial f_x: \RR^m \to \RR^n$;
\item $\reindex{\partial g}{f}_x = \partial g_{f(x)}: \RR^n \to \RR^k$
\end{itemize}

\noindent The composition $(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp
\partial f): (\RR^m, \const(\RR^m)) \to (\RR^k, \const(\RR^k))$ corresponds exactly to the chain rule, since
$(\reindex{\partial g}{f} \comp \partial f)_x = {\partial g}_{f(x)} \comp {\partial f}_x$.

@@

Now recall that the category $\Fam(\LatGal)$ is the Grothendieck construction $\Grothendieck{X}\Fam(X,
\LatGal)$, with:
\begin{itemize}
\item as objects $(X, \partial X)$ all pairs of a set $X$ and and for each $x \in X$, a bounded lattice
$\partial X_x$;
\item as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of functions $f: X \to
Y$ and Galois connections $\partial f: \partial X_x \to \partial Y_{f(x)}$.
\end{itemize}

>>>>

\subsubsection{Forward mode automatic differentiation}

\begin{definition}
Define the functor $\tangents$ which sends every vector space $\RR^n$ to itself and every differentiable map
$f: \RR^m \to \RR^n$ to the function $\tangents(f) = x \mapsto (f(x), \pushf{f}_x): \RR^m \to \RR^n \times
(\RR^m \linearto \RR^n)$, associating to every point its image in $f$ and the forward derivative of $f$ at
that point.
\end{definition}

@@

\subsubsection{$\Diff$ as a subcategory of $\Fam(\FinVect)$}

Recall that $\Fam(\FinVect)$ denotes the Grothendieck construction $\Grothendieck{X}\Fam(X, \FinVect)$. The
functor $\Diff \hookrightarrow \Fam(\FinVect)$ which send objects $\RR^n$ to constant families of tangent
spaces $(\RR^n, \const(\RR^n))$ and morphisms $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ to pairs
$(f_1, f_2)$ of differentiable functions $f_1: \RR^m \to \RR^n$ and families of linear maps $f_2: \{{f_2}_x:
\RR^m \linearto \RR^n\}_{x \in \RR^m}$ exhibits $\Diff$ as equivalent to a full subcategory of
$\Fam(\FinVect)$.

For any $(f, \partial f): (\RR^m, \const(\RR^m)) \to (\RR^n, \const(\RR^n))$ and $(g, \partial g): (\RR^n,
\const(\RR^n)) \to (\RR^k, \const(\RR^k))$ in that subcategory, we have that for any $x \in \RR^m$:
\begin{itemize}
\item $\partial f_x: \RR^m \to \RR^n$;
\item $\reindex{\partial g}{f}_x = \partial g_{f(x)}: \RR^n \to \RR^k$.
\end{itemize}

\noindent The composition $(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp
\partial f): (\RR^m, \const(\RR^m)) \to (\RR^k, \const(\RR^k))$ corresponds exactly to the forward chain rule,
since $(\reindex{\partial g}{f} \comp \partial f)_x = {\partial g}_{f(x)} \comp {\partial f}_x$.

\subsubsection{Reverse mode automatic differentiation}

\begin{definition}
Define the functor $\cotangents$ which sends every vector space $\RR^n$ to itself and every differentiable map
$f: \RR^m \to \RR^n$ to the function $\cotangents(f) = x \mapsto (f(x), \pullf{f}_x): \RR^m \to \RR^n \times
(\RR^n \linearto \RR^m)$.
\end{definition}

\noindent and again we can define a corresponding category $\Diff^*$ of such maps where composition of the
cotangent maps respects the backward chain rule, with in this case $\Diff^*$ equivalent to a full subcategory
of $\Fam(\FinVect^\op)$.

@@

Now recall that $\Fam(\LatGal)$ is the Grothendieck construction $\Grothendieck{X}\Fam(X, \LatGal)$, with:
\begin{itemize}
\item as objects $(X, \partial X)$ all pairs of a set $X$ and and for each $x \in X$, a bounded lattice
$\partial X_x$;
\item as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of functions $f: X \to
Y$ and Galois connections $\partial f: \partial X_x \to \partial Y_{f(x)}$.
\end{itemize}
